\renewcommand{\abstractname}{Acknowledgements}
\begin{abstract}
  There are several people I'd like to thank for making this thesis possible. First
  and foremost, my advisor Prof. Sanjit Seshia, who encouraged me to pursue
  my own research ideas, and was constantly supportive
  and patient with the many setbacks in my eventually
  doomed graduate career. I'd also like to thank Prof. Bill Harris, Prof. Somesh Jha, and
  Drew Davidson, who have had a major role to play in defining and concretizing the
  ideas presented in this thesis. It is only due to our sustained collaboration that this
  work managed to see the light of day. Also thanks to Rohit Sinha,
  who was heavily involved in the conception and inital work on these ideas.
  I would also like to thank Prof. Bj{\"o}rn Hartmann, whose valuable comments made this
  thesis much more robust and well-rounded. I'd also like to thank him and Sanjit for being
  considerate about the difficult circumstances surrounding the conclusion of this work.

  Before I thank others, I must describe how working
  on this thesis (and graduate school) affected me personally. I will forever remember
  this time as one of the darkest periods of my life. It was partly due to severe
  personal difficulties that led to chronic and often self-destructive
  depression. It was a constant
  hindrance towards my ability to make progress with research. It made a notable dent
  in relationships with friends and family, but I'd like to thank those who stood by
  me. First of all, I'd like to thank my girlfriend Vasuki, who
  has had to bear the most immediate impact of this dark period. Despite her own
  issues, she was often the only one who truly listened to me, and made a conscious
  effort to understand my problems and help me in concrete ways. A lot of the
  motivation to finish this work and move on in life wouldn't have been possible
  without her. I'd also like to thank my parents, who have
  always supported me in everything I chose
  to do, including my successes and failures, and positively encouraged me to
  never underestimate my abilities.
  I must thank my other friends in Berkeley -- Avinash, Vaishnavi, Debanjan, Sharanya,
  and Vikram who provided me with the sense of community that invariably played
  a large role in surviving this painful phase. Avinash in particular
  was a constantly calming
  presence in the crunch period when I missed the final
  submission deadline and was facing severe consequences. I'd also like to thank
  a few of my other friends -- Anasuya, Abhay, Pradeep, Pramod, Vishal, Namit, and
  Ravi, who despite being physically further away, understood my pain and
  frustration. My conversations with Anasuya in the worst of times
  have been an invaluable source of stability.

  Finally, I'd like to thank the extremely helpful people at the Disabled Students Program
  (DSP) in Berkeley and my therapist, without whose help I'd have
  had to pay thousands of dollars to submit the thesis.
\end{abstract}