\label{ch:intro}

In the context of hardware and software systems, formal verification is the act of
proving or disproving the correctness of intended algorithms underlying a system with
respect to a certain formal specification or property, using formal methods of
mathematics. Formal verification can be helpful in proving the correctness of systems
such as: cryptographic protocols, combinational circuits, digital circuits with internal
memory, and software expressed as source code. Verification uses a wide variety of
techniques, in particular logic calculi, formal languages, automata theory, and program
semantics, but also type systems and algebraic data types to problems in software and
hardware specification and verification.

The verification of these systems is done by providing a formal proof on an abstract
mathematical model of the system, the correspondence between the mathematical model and
the nature of the system being otherwise known by construction. Examples of mathematical
objects often used to model systems are: finite state machines, labeled transition
systems, Petri nets, vector addition systems, timed automata, hybrid automata, process
algebra, formal semantics of programming languages such as operational semantics,
denotational semantics, axiomatic semantics and Hoare logic.

\section{Software Verification}

The goal of software verification is to assure that software fully satisfies all the
expected requirements. Software verification has been successful for improving the
quality of computer programs. Several fundamental concepts were invented in the last
decade which made it possible to scale the technology from tiny examples to real
programs. Predicate abstraction \cite{ball01} with counterexample guided abstraction
refinement (CEGAR) \cite{clarke03} and lazy abstraction \cite{henzinger02} is one such
set of techniques. Lazy abstraction with Interpolants \cite{mcmillan06} is another
approach. Several of these approaches are flexible and can be specialized or extended for
sub-classes of programs or specifications.

\section{Verifying Heap-manipulating Programs}
A heap-manipulating program is one that updates heap memory using low-level memory
operations (such as allocating or deallocating memory, or modifying pointers). Dealing
with programs with pointers and dynamic linked data structures is among the most
challenging tasks of formal analysis and verification due to a need to deal with infinite
sets of reachable program configurations having the form of complex graphs. This task
becomes even more complicated when considering low-level memory operations such as
pointer arithmetic, safe usage of pointers with invalid targets, block operations with
memory, reinterpretation of the memory contents, or address alignment. Despite the rapid
progress in the area of formal program analysis and verification, fully automated
approaches capable of efficiently handling sufficiently general classes of dynamic
linked data structures in the form used in low-level code are still missing.

\section{Oracle-guided Heap Interpolant Synthesis}
Traditional formal methods do not scale to the size of software found in modern computer
systems. This is further exacerbated in the case of heap-manipulating programs.
Verification also currently requires highly specialized engineers with deep knowledge of
software technology and mathematical theorem-proving techniques. These constraints make
current formal verification techniques expensive and time-consuming.

The formal verification community has relentlessly pushed for automation, and it is
indeed the right strategy for many problems. But for many other problems, human insight
and involvement remain crucial. For instance, the steps of writing a specification,
typically in the form of properties (assertions), creating an environment model,
typically as input constraints or a state machine currently both require significant
human intervention. Even the task of running the verifier, such as a model checker, which
is usually thought of as a ``push-button'' step might require human insight such as hints
to the verifier in the form of suitable abstraction techniques or inductive invariants.
If the verifier returns a counterexample trace, then analysing it and finally fixing the
design to debug the program also needs human input. Despite the focus on automating the
verification process, we continue to need human insight in a variety of tasks, including
writing specifications, creating models, guiding the verification engine, debugging and
error localization, and repair.

It might be a while before we have verification systems that don't require any human
expertise, but an alternative approach to consider is to change the way humans provide
insight to the verifier. Today, mostly domain experts interact with verification tools,
but such experts are few and expensive. Besides, even experts might sometimes struggle to
make progress at different stages of the algorithm. We believe that the experts and
automated tools can be assisted in the verification process by a crowd of non-expert
humans performing relatively simpler tasks. Each task might involve pattern recognition
or other cognitive operations that humans are typically good at. The main challenges are
to recognise steps in the verification process where human insight is useful, find ways
to transform these steps into tasks that non-expert humans can perform, and finally
combine the results of these tasks to use in the verification process.

Along these lines, our work describes \verifier, an Oracle-guided
verification algorithm for proving the safety of heap-manipulating programs. We present
an extension of the interpolant-based verification algorithm in \cite{mcmillan06} to work
for heaps, along with an interface that collects insight from an external Oracle. In our
case, the Oracle is a human user (although it could be any other Oracle that can interact
with the verifier). The user plays a graphical game that abstracts away details of the
verification algorithm, allowing non-experts to play. Essentially, our system is a
verification tool for heap-manipulating programs where crucial input is collected from
human users.

\section{Outline}

The rest of this thesis is organized as follows. \autoref{ch:overview} gives a high-level
overview of our approach with an example, laying an intuitive foundation for the formal
description that follows. \autoref{ch:background} presents an extensive summary of the
framework for finding inductive invariants from interpolants, and the Impact algorithm
from \cite{mcmillan06}. \autoref{ch:heap-patterns} describes our formalism for
representing heaps and heap patterns, and modeling heap-manipulating programs for
verification. \autoref{ch:heap-impact-algorithm} gives a detailed description of our
main algorithm \verifier, as an extension of the Impact algorithm. Our algorithm
requires an Oracle, which we have chosen to be a human user, and
\autoref{ch:interface-oracle} presents our user interface with some analysis.
\autoref{ch:conclusion} discusses related work, and concludes.

\subsubsection{Collaboration}
The work in this thesis is part of the collaboration with Somesh
Jha\footnote{University of Wisconsin, Madison}, William (Bill)
Harris\footnote{Georgia Institute of Technology}, and Drew
Davidson\footnote{University of Wisconsin, Madison}. Somesh and Bill were immensely
insightful in helping narrow down the problem domain.
The heap pattern formalism that forms a major part of the formal background for this
thesis was conceived over the course of several regular weekly meetings with Bill, Somesh,
and Sanjit. This involved understanding and analyzing several existing approaches, and
narrowing down on the properties we cared about for an oracle-guided framework. They
were also instrumental in advancing theoretical analysis of heap patterns
(this is still work in progress, and didn't make it into the thesis).

Drew and I worked together implementing our ideas, initially building on top
of Predator \cite{dudka13}, and later in \cpachecker \cite{cpachecker}. Later on,
I worked on finishing this implementation, and also independently built the
user interface to human oracles \autoref{ch:interface-oracle}.

This work would not have been possible without our regular weekly discussions, constant
guidance from Bill, Somesh, and Sanjit, and the shared effort with Drew on implementing
these ideas.

